Finally Tagless Staged Interpreters for Simpler Typed Languages

### 2007-04-06 tagless/tagless-final.tex
Shortening

\titlebanner{banner above paper title}        % These are ignored unless
\preprintfooter{short description of paper}   % 'preprint' option specified.

\title{Staged typed type-preserving tagless ``interpreters'', \emph{finally}}
\subtitle{Subtitle Text, if any}

@@

Our code ensures statically that an object term is well-typed and
closed, yet is surprisingly simple and obvious in hindsight.  We stress
that it has been an open problem to
interpret a typed object language in a typed metalanguage without
tagging or type\hyp system extensions.  For example, \citet{taha-tag}
state that ``expressing such an interpreter in a statically typed
programming language is a rather subtle matter. In fact, it is only
recently that some work on programming type-indexed values in ML
\citep{yang-encoding} has given a hint of how such a function can be
expressed.''  We further discuss related work in~\S\ref{related}.

Our contributions are as follows.
\begin{enumerate*}
\item We build interpreters that evaluate (in direct or continuation\hyp passing styles),
  compile, and partially evaluate a typed higher-order object language
   in a typed metalanguage.
\item All these interpreters use no type tags, patently never get stuck,
    and need no advanced type-system features such as GADTs, dependent types,
    or intentional type analysis.
\item The partial evaluator avoids polymorphic lift and delays binding-time
    analysis.
\item We use the type system of the metalanguage
    to check statically for well-typed and closed object programs.
\item We show clean, comparable implementations in MetaOCaml and Haskell.
\item We specify a functor signature that encompasses all our interpreters, from
    evaluation and compilation to partial evaluation and CPS transformation.
\item We point a clear way to extend the object language with more features
  such as state.
\item We describe an approach to self\hyp interpretation compatible with the
  above.  Self\hyp interpretation turned out harder than expected.
\end{enumerate*}

The structure of our paper is as follows. In~\S\ref{language}, we
define the OCaml and Haskell interfaces of the interpreter for our
language and describe two instances, evaluating a term to a
metalanguage value and computing term's size.
In~\S\ref{compiler}, we stage the evaluator, obtaining the tagless
compiler. \S\ref{PE} describes partial evaluation, in
particular, the problem posed by the need to delay the 
binding-time analysis of term abstractions and the apparent 
need for a typecase. We then show two solutions. The first, imperfect
one, uses GADT. The better and more portable solution bakes in the
typecase idiom into the interpreter interface, eliminating the need for
advanced type systems and permitting implementations in Haskell98 or OCaml.
The discussion \S\ref{discussion} presents, inter alia, CBN CPS
interpreter and the CBV CPS transformer. We then describe 
self-interpretation (\S\ref{selfinterp}),
discuss related work (\S\ref{related}) and then conclude.

@@

Given this implementation of~|P|, our running example
\begin{code}
let module E = EX(P) in E.test1 ()
\end{code}
evaluates to
\begin{code}
{P.st = Some true; P.dy = .<true>.}
\end{code}
of type |('a, bool, bool) P.repr|.  Unlike with~|C| in~\S\ref{compiler},
a $\beta$-reduction has been statically performed to yield |true|.  More
interestingly, whereas |testpowfix7| compiles to a code value with many
$\beta$-redexes in~\S\ref{compiler}, the partial evaluation
\begin{code}
let module E = EX(P) in E.testpowfix7
\end{code}
gives
\begin{code}
{P.st = Some <fun>;
 P.dy = .<fun x_1 -> x_1 * (x_1 * (x_1 * (x_1 *
                    (x_1 * (x_1 * x_1)))))>.}
\end{code}
as desired.

@@

The object language generally inherits the evaluation strategy from
the metalanguage. Evaluating an application |e1 e2| in OCaml requires
the evaluation of |e2| before performing the application. Therefore,
when evaluating the encoding of the object language application 
|app e1 e2|, |e2| will always have to be evaluated. One may wonder if it is
possible to represent a call-by-name (CBN) object language in a
call-by-value (CBV) metalanguage. The answer, given already by 
\citet{reynolds-definitional,reynolds-relation} and \citet{PlotkinCBN}, is yes thanks to CPS, which they
specifically introduced 
to make the evaluation strategy of a definitional interpreter
independent of the strategy of the metalanguage. We demonstrate the
same independence in the typed setting \oleg{mention the quote from
Reynolds that typed setting would be nice?}\jacques{I think so, but from
which paper? The one ReynoldsPlotkin paper is typed}\ccshan{I think Oleg
meant ``It should also be possible to define languages with a highly
 refined syntactic type structure. Ideally, such a treatment should be
 metacircular, in
 the sense that the type structure used in the defined language should be adequate for the
 defining language'' \citep{reynolds-definitional}, but perhaps Jacques
 meant another paper by ``ReynoldsPlotkin''?}. Specifically, we
demonstrate CBN evaluation for our object language in OCaml, which is
call-by-value. We do so by building a CBN CPS interpreter for the object
language.

@@

\begin{code}
module RCN = struct
  type ('c,'sv,'dv) repr = 
     {ko: 'w. ('sv -> 'w) -> 'w}
  let int (x:int) = {ko = fun k -> k x}
  let add e1 e2 = 
    {ko = fun k -> e1.ko (fun v1 -> 
                   e2.ko (fun v2 -> k (v1+v2)))}
  let if_ eb et ee = 
    {ko = fun k -> eb.ko 
         (fun vb -> if vb then (et ()).ko k 
                    else (ee ()).ko k)}
  ...
  let lam f = {ko = fun k -> k f}
  let app e1 e2 = 
     {ko = fun k -> e1.ko (fun f -> (f e2).ko k)}
  let fix f = 
     let rec fx f n = app (f (lam (fx f))) n in 
     lam(fx f)

  let get_res x = x.ko (fun v -> v)
end
\end{code}

@@

Because |RCN| has the signature |Symantics|, we can instantiate our previous
examples with it, and all works as expected.  More interesting
is the example $(\fun{x}1)\bigl((\fix{f}f)\mathinner2\bigr)$, which terminates
under CBN but not CBV\@.
\begin{code}
module EXS(S: Symantics) = struct
 open S
 let diverg () = 
   app (lam (fun x -> int 1)) 
       (app (fix (fun f -> f)) (int 2))
end
\end{code}

If we interpret |EXS| with the |R| interpreter of
\S\ref{S:interpreter-RL} realized in OCaml
\begin{code}
let module M = EXS(R) in M.diverg ()
\end{code}
we never get any result because the above diverges. In contrast, if we use
the CBN interpreter
\begin{code}
let module M = EXS(RCN) in RCN.get_res (M.diverg ())
\end{code}
we get the result |1|.

\oleg{mention some formal results? Like indifference theorems by
  Plotkin? And say we generalized their result to the type setting?
  Mention this in the conclusions?}
\jacques{you have strayed outside what I know -- what Plotkin results?
  I always consider mentionning formal results a good thing}

@@

We can easily change our CBN CPS interpreter into a CBV CPS
interpreter, by replacing only one definition:
\begin{code}
module RCV = struct
  include RCN
  let lam f = {ko = 
    fun k -> k (fun e -> e.ko 
         (fun v -> f {ko = fun k -> k v}))}
end
\end{code}
Now when we apply an abstraction to an argument, we always force
the evaluation of that argument before proceeding. This approach is in
line with \citet{reynolds-relation} (albeit in a typed setting). The interpreter RCV
is useful for achieving CBV evaluation of the object language
no matter whether the metalanguage is CBV or CBN.

@@

It is instructive to see the result of evaluating a few examples. With 
|CPST| instantiated with a straightforward compiler |C| as above, our
running test |T.test1 ()| gives
\begin{code}
.<fun x_5 -> ((fun x_2 -> 
        (x_2 (fun x_3 -> fun x_4 -> (x_4 x_3))))
        (fun x_6 -> ((fun x_1 -> (x_1 (true))) 
        (fun x_7 -> ((x_6 x_7) x_5)))))>.
\end{code}

@@

\subsection{State and imperative features}

@@

Instead of adding higher-rank and higher-kind polymorphism to our object language (along
with polymorphism over kinds), we add let-bound polymorphism.  As usual,
we can add a new typing rule
\begin{equation*}
    \begin{prooftree}
        e_1:t_1 \quad \subst{e_2}{x}{e_1}:t_2
        \justifies \be{x=e_1} e_2 : t_2
    \end{prooftree}
    .
\end{equation*}
The pre-encoding of a let\hyp expression is trivial.
\begin{align*}
    \encode{\be{x=e_1}e_2} \quad &= \quad \be{x=\encode{e_1}} \encode{e_2}
\intertext{A \emph{context} is an object term with a hole~$[~]$.  The
hole may occur under a binder, so plugging a term into the context may
capture free variables of the term.  By pre-encoding a hole to a hole,
we extend pre-encoding from a translation on terms to a translation on
contexts.}
    \encode{[~]} \quad &= \quad [~]
\end{align*}
We define an interpreter in the object language to be not a term but
a context.  For example, the evaluator is the context
\begin{align*}
    \si[~] &=
    \begin{tabular}[t]{@{}Ml@{}>{{}}Ml@{}Ml@{}}
        \be{\ident{int} &= \fun{x} x&} \\
        \be{\ident{add} &= \fun{x} \fun{y} x+y&} \\
        \be{\ident{if\_}&= \fun{b} \fun{t} \fun{e} \cond{b}{t\,0}{e\,0}&} \\
        \be{\ident{lam} &= \fun{f} f&} \\
        \be{\ident{app} &= \fun{f} \fun{x} fx&} \\
        \be{\ident{fix} &= \fun{g} \fix{f} \fun{x} gfx&} [~],
    \end{tabular}
\intertext{and the size\hyp measurer is the context}
    \mathrm{SZ}[~] &=
    \begin{tabular}[t]{@{}Ml@{}>{{}}Ml@{}Ml@{}}
        \be{\ident{int} &= \fun{x} 1&} \\
        \be{\ident{add} &= \fun{x} \fun{y} x+y+1&} \\
        \be{\ident{if\_}&= b + t\,0 + e\,0&} \\
        \be{\ident{lam} &= \fun{f} f\,0 + 1&} \\
        \be{\ident{app} &= \fun{f} \fun{x} f + x + 1&} \\
        \be{\ident{fix} &= \fun{g} g\,0 + 1&} [~].
    \end{tabular}
\end{align*}
To interpret an object term~$e$ using an interpreter $I[~]$ is to
evaluate the object term~$I[\encode{e}]$.  $\si$ is
a self\hyp interpreter in the following sense.
\begin{prop}
    $\si[\encode{e}]$ is observationally equivalent to~$e$.
\end{prop}
As a corollary, we can pre-encode the self\hyp interpreter itself as
a context: the term $\si[\encode{\si[\encode{e}]}]$ is observationally
equivalent to $\si[\encode{e}]$, and in turn to~$e$.  In other words,
$\si$ can interpret itself.

@@

From reading \citet{WalidICFP02,taha-tag,xi-guarded,peyton-jones-simple}, our 
impression is that the claim is:
\begin{enumerate}
\item One needs to encode a typed language in a typed language based on
a sum type (at some level of the hierarchy),
\item It is not possible to write a \emph{direct} interpreter 
for such an encoding of a typed language
in a typed language, without either using a
very advanced type system or incurring tagging/untagging overhead.
\item Thus an indirect interpreter is necessary, which needs a universal
  type (and hence tagging),
\item Thus any self-interpreter must have tags.
\item Hence the self-interpreter cannot be Jones-optimal.
\end{enumerate}
While the logic is sound, we showed that the premise in the very first step
was not valid.

In particular, \citet[\S5]{taha-tag}
states that ``expressing such an interpreter in a statically typed
programming language is a rather subtle matter. In fact, it is only
recently that some work on programming type-indexed values in ML
\citep{yang-encoding} has given a hint of how such a function can be
expressed.''

>>>>

\makeatletter
\titlebanner{Submitted to \@conferencename}        % These are ignored unless
\preprintfooter{\rlap{\@titletext}}   % 'preprint' option specified.
\def \@formatyear {\llap{\number\year/\number\month/\number\day}}
\makeatother

\title{Finally Tagless Staged Interpreters for Simpler Typed Languages}
%Staged typed type-preserving tagless ``interpreters'', \emph{finally}
%Finally tagless, simply typed, partially evaluated
\subtitle{\raisebox{-2.5em}[0pt][0pt]{\normalsize\it\begin{minipage}{30em}
    It should also be possible to define languages with a highly
    refined syntactic type structure. Ideally, such a treatment should
    be metacircular, in the sense that the type structure used in the
    defined language should be adequate for the defining language.
    \hfill\rm\citep{reynolds-definitional}
\end{minipage}}}

@@

Our contributions are as follows.
\begin{enumerate*}
\item We build interpreters that evaluate (\S\ref{language}),
    compile (or evaluate with staging) (\S\ref{compiler}), and partially evaluate (\S\ref{PE}) a typed higher-order object language
   in a typed metalanguage, in direct and continuation\hyp passing styles
   (\S\ref{discussion}).
\item All these interpreters use no type tags, patently never get stuck,
    and need no advanced type-system features such as GADTs, dependent types,
    or intentional type analysis.
\item The partial evaluator avoids polymorphic lift and delays binding-time
    analysis.  It bakes a type-to-type map into the interpreter
    interface to eliminate the need for GADTs and thus remain portable
    across Haskell 98 and ML.
\item We use the type system of the metalanguage
    to check statically that an object program is well-typed and closed.
\item We show clean, comparable implementations in MetaOCaml and Haskell.
\item We specify a functor signature that encompasses all our interpreters, from
    evaluation and compilation (\S\ref{language}) to partial evaluation and CPS transformation (\S\ref{PE}).
\item We point a clear way to extend the object language with more features
    such as state (\S\ref{state}).
\item We describe an approach to self\hyp interpretation compatible with the
  above (\S\ref{selfinterp}).  Self\hyp interpretation turned out harder than expected.
\end{enumerate*}
Our code is surprisingly simple and obvious in hindsight, but
it has been an open problem to
interpret a typed object language in a typed metalanguage without
tagging or type\hyp system extensions.  For example, \citet{taha-tag}
state that ``expressing such an interpreter in a statically typed
programming language is a rather subtle matter. In fact, it is only
recently that some work on programming type-indexed values in ML
\citep{yang-encoding} has given a hint of how such a function can be
expressed.''  We further discuss related work in~\S\ref{related}.

@@

Given this implementation of~|P|, our running example
\begin{code}
let module E = EX(P) in E.test1 ()
\end{code}
evaluates to
\begin{code}
{P.st = Some true; P.dy = .<true>.}
\end{code}
of type |('a, bool, bool) P.repr|.  Unlike with~|C| in~\S\ref{compiler},
a $\beta$-reduction has been statically performed to yield |true|.  More
interestingly, whereas |testpowfix7| compiles to a code value with many
$\beta$-redexes in~\S\ref{compiler}, the partial evaluation
\begin{code}
let module E = EX(P) in E.testpowfix7
\end{code}
gives the desired result
\begin{code}
{P.st = Some <fun>;
 P.dy = .<fun x_1 -> x_1 * (x_1 * (x_1 * (x_1 *
                    (x_1 * (x_1 * x_1)))))>.}
\end{code}

@@

The object language generally inherits the evaluation strategy from
the metalanguage. Evaluating an application |e1 e2| in OCaml requires
the evaluation of |e2| before performing the application. Therefore,
when evaluating the encoding of the object language application 
|app e1 e2|, |e2| will always have to be evaluated. One may wonder if it is
possible to represent a call-by-name (CBN) object language in a
call-by-value (CBV) metalanguage. The answer, given already by 
\citet{reynolds-definitional,reynolds-relation} and \citet{PlotkinCBN}, is yes thanks to CPS, which they
specifically introduced 
to make the evaluation strategy of a definitional interpreter
independent of the strategy of the metalanguage. We demonstrate the
same independence in the typed setting. Specifically, we
demonstrate CBN evaluation for our object language in OCaml, which is
call-by-value. We do so by building a CBN CPS interpreter for the object
language.

@@

\begin{code}
module RCN = struct
  type ('c,'sv,'dv) repr = 
    {ko: 'w. ('sv -> 'w) -> 'w}
  let int (x:int) = {ko = fun k -> k x}
  let add e1 e2 = 
    {ko = fun k -> e1.ko (fun v1 -> 
                   e2.ko (fun v2 -> k (v1+v2)))}
  let if_ eb et ee = 
    {ko = fun k -> eb.ko 
         (fun vb -> if vb then (et ()).ko k 
                          else (ee ()).ko k)}
  let lam f = {ko = fun k -> k f}
  let app e1 e2 = 
    {ko = fun k -> e1.ko (fun f -> (f e2).ko k)}
  let fix f = 
    let rec fx f n = app (f (lam (fx f))) n in 
    lam (fx f)
  let get_res x = x.ko (fun v -> v)
end
\end{code}

@@

Because |RCN| has the signature |Symantics|, we can instantiate our previous
examples with it, and all works as expected.  More interesting
is the example $(\fun{x}1)\bigl((\fix{f}f)\mathinner2\bigr)$, which terminates
under CBN but not CBV\@.
\begin{code}
module EXS(S: Symantics) = struct
 open S
 let diverg () = app (lam (fun x -> int 1)) 
                     (app (fix (fun f -> f)) (int 2))
end
\end{code}
If we interpret |EXS| with the |R| interpreter of
\S\ref{S:interpreter-RL}
\begin{code}
let module M = EXS(R) in M.diverg ()
\end{code}
we never get any result because the above diverges. In contrast, if we use
the CBN interpreter
\begin{code}
let module M = EXS(RCN) in RCN.get_res (M.diverg ())
\end{code}
we get the result |1|.

@@

We can easily change our CBN CPS interpreter into a CBV CPS
interpreter, by replacing only one definition:
\begin{code}
module RCV = struct
  include RCN
  let lam f = {ko = fun k -> k (fun e ->
    e.ko (fun v -> f {ko = fun k -> k v}))}
end
\end{code}
Now when we apply an abstraction to an argument, we always force
the evaluation of that argument before proceeding. This approach is in
line with \citet{reynolds-relation} (albeit in a typed setting). The interpreter RCV
is useful for achieving CBV evaluation of the object language
whether the metalanguage is CBV or CBN.

@@

It is instructive to evaluate a few examples. With 
|CPST| instantiated with the compiler~|C| above, our
running test |T.test1 ()| gives
\begin{code}
.<fun x_5 -> ((fun x_2 -> 
        (x_2 (fun x_3 -> fun x_4 -> (x_4 x_3))))
        (fun x_6 -> ((fun x_1 -> (x_1 (true))) 
        (fun x_7 -> ((x_6 x_7) x_5)))))>.
\end{code}

@@

\subsection{State and imperative features}
\label{state}

@@

Instead of adding higher-rank and higher-kind polymorphism to our object language (along
with polymorphism over kinds), we add let-bound polymorphism.  As usual,
we can add a new typing rule
\begin{equation*}
    \begin{prooftree}
        e_1:t_1 \quad \subst{e_2}{x}{e_1}:t_2
        \justifies \be{x=e_1} e_2 : t_2
    \end{prooftree}
    .
\end{equation*}
The pre-encoding of a let\hyp expression is trivial.
\begin{align*}
    \encode{\be{x=e_1}e_2} \quad &= \quad \be{x=\encode{e_1}} \encode{e_2}
\intertext{A \emph{context} is an object term with a hole~$[~]$.  The
hole may occur under a binder, so plugging a term into the context may
capture free variables of the term.  By pre-encoding a hole to a hole,
we extend pre-encoding from a translation on terms to one on
contexts.}
    \encode{[~]} \quad &= \quad [~]
\end{align*}
We define an interpreter in the object language to be not a term but
a context.  For example, the evaluator is the context
\begin{align*}
    \si[~] &=
    \begin{tabular}[t]{@{}Ml@{}>{{}}Ml@{}Ml@{}}
        \be{\ident{int} &= \fun{x} x&} \\
        \be{\ident{add} &= \fun{x} \fun{y} x+y&} \\
        \be{\ident{if\_}&= \fun{b} \fun{t} \fun{e} \cond{b}{t\,0}{e\,0}&} \\
        \be{\ident{lam} &= \fun{f} f&} \\
        \be{\ident{app} &= \fun{f} \fun{x} fx&} \\
        \be{\ident{fix} &= \fun{g} \fix{f} \fun{x} gfx&} [~],
    \end{tabular}
\intertext{and the size\hyp measurer is the context}
    \mathrm{SZ}[~] &=
    \begin{tabular}[t]{@{}Ml@{}>{{}}Ml@{}Ml@{}}
        \be{\ident{int} &= \fun{x} 1&} \\
        \be{\ident{add} &= \fun{x} \fun{y} x+y+1&} \\
        \be{\ident{if\_}&= b + t\,0 + e\,0&} \\
        \be{\ident{lam} &= \fun{f} f\,0 + 1&} \\
        \be{\ident{app} &= \fun{f} \fun{x} f + x + 1&} \\
        \be{\ident{fix} &= \fun{g} g\,0 + 1&} [~].
    \end{tabular}
\end{align*}
To interpret an object term~$e$ using an interpreter $I[~]$ is to
evaluate the object term~$I[\encode{e}]$.  $\si$ is
a self\hyp interpreter in the following sense.
\begin{prop}
    $\si[\encode{e}]$ is observationally equivalent to~$e$.
\end{prop}
As a corollary, we can pre-encode the self\hyp interpreter itself as
a context: the term $\si[\encode{\si[\encode{e}]}]$ is observationally
equivalent to $\si[\encode{e}]$, and in turn to~$e$.  In other words,
$\si$ can interpret itself.

@@

\Citet{WalidICFP02}, \citet{taha-tag}, \citet{xi-guarded}, and
\citet{peyton-jones-simple} seem to argue as follows that a self\hyp
interpreter of a typed language cannot be Jones\hyp optimal.
\begin{enumerate}
\item One needs to encode a typed language in a typed language based on
a sum type (at some level of the hierarchy),
\item It is not possible to write a \emph{direct} interpreter 
for such an encoding of a typed language
in a typed language, without either using a
very advanced type system or incurring tagging/untagging overhead.
\item Thus an indirect interpreter is necessary, which needs a universal
  type (and hence tagging),
\item Thus any self-interpreter must have tags.
\item Hence the self-interpreter cannot be Jones-optimal.
\end{enumerate}
While the logic is sound, we showed that the premise in the very first step
was not valid.
